---
title: "Basic examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{basic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, message=FALSE}
library(blendR)
```

```{r, warning=FALSE, message=FALSE}
library(survHE)
library(INLA)
library(sn)
```


## INLA piece-wise exponential and external knowledge

First load the package data and take a look.

```{r}
data("TA174_FCR", package = "blendR")
head(dat_FCR)
```

Next, fit the observed data with a piece-wise exponential model using INLA.

```{r}
## observed estimate
obs_Surv <- surv_est_inla(data = dat_FCR,
                          cutpoints = seq(0, 180, by = 5))
```

For the external model first we create some synthetic data consistent with user-defined constraints as follows.

```{r}
## external estimate
data_sim <- ext_surv_sim(t_info = 144,
                         S_info = 0.05,
                         T_max = 180)
```

Then fit a gompertz model to this.

```{r}
ext_Surv <- fit.models(formula = Surv(time, event) ~ 1,
                       data = data_sim,
                       distr = "gompertz",
                       method = "hmc",
                       priors = list(gom = list(a_alpha = 0.1,
                                                b_alpha = 0.1)))
```

Lastly, we can run the blending step.

```{r}
blend_interv <- list(min = 48, max = 150)
beta_params <- list(alpha = 3, beta = 3)

ble_Surv <- blendsurv(obs_Surv, ext_Surv, blend_interv, beta_params)

plot(ble_Surv)
```

## Two HMC survHE fits

In the same way as above we fit two survival curves and then blend them together.
In this case we use two exponential curves using the survHE package to fit them.

```{r}
obs_Surv2 <- fit.models(formula = Surv(death_t, death) ~ 1,
                        data = dat_FCR,
                        distr = "exponential",
                        method = "hmc")

ext_Surv2 <- fit.models(formula = Surv(time, event) ~ 1,
                       data = data_sim,
                       distr = "exponential",
                       method = "hmc")
```

```{r}
ble_Surv2 <- blendsurv(obs_Surv2, ext_Surv2, blend_interv, beta_params)

# kaplan-meier
km <- survfit(Surv(death_t, death) ~ 1, data = dat_FCR)

plot(ble_Surv2) +
  geom_line(aes(km$time, km$surv, colour = "Kaplan-Meier"),
            size = 1.25, linetype = "dashed")
```

## flexsurv frequentist background fit

The last example is for an HMC and frequentist survival model using the flexsurv package directly.

```{r}
obs_Surv3 <- fit.models(formula = Surv(death_t, death) ~ 1,
                        data = dat_FCR,
                        distr = "exponential",
                        method = "hmc")

ext_Surv3 <- flexsurv::flexsurvreg(formula = Surv(time, event) ~ 1,
                       data = data_sim,
                       dist = "gompertz")

ble_Surv3 <- blendsurv(obs_Surv3, ext_Surv3, blend_interv, beta_params)

plot(ble_Surv3)
```
