---
title: "{expertsurv} package examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{{expertsurv} package examples}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, message=FALSE}
library(blendR)
```


```{r install-pkgs, warning=FALSE, message=FALSE, eval=TRUE, echo=FALSE}
if (!requireNamespace("survHEhmc")) remotes::install_github('giabaio/survHEhmc')
```

```{r, warning=FALSE, message=FALSE}
library(survHE)
library(survHEhmc)
library(INLA)
```

```{r limit-cores, echo=FALSE}
# R CMD check only allows a maximum of two cores
options(mc.cores=2)
inla.setOption(num.threads = 2)
```

## `{expertsurv}` background fit

The next example is using the [`{expertsurv}`](https://github.com/Philip-Cooney/expertsurv) package to fit the external model.

We can create the external survival curve using the penalised regression model.

```{r eval=FALSE}
# if (!requireNamespace("expertsurv", quietly = TRUE)) 
#   remotes::install_github("Philip-Cooney/expertsurv")

library(expertsurv)

param_expert <- list()

# S = 0.05
param_expert[[1]] <- data.frame(dist = "norm",
                                wi = 1,
                                param1 = 0.2,
                                param2 = 0.05,
                                param3 = NA)

# S = 0
param_expert[[2]] <- data.frame(dist = "norm",
                                wi = 1,
                                param1 = 0.05,
                                param2 = 0.005,
                                param3 = NA)

timepoint_expert <- c(100, 180)

# dummy data
data <- data.frame(time =50, event = 0)

# don't provide any data so its all based on the prior
ext_Surv <- fit.models.expert(formula = Surv(time,event) ~ 1,
                              data = data,
                              distr = "gomp",
                              method = "hmc",
                              iter = 1000,
                              opinion_type = "survival",
                              times_expert = timepoint_expert, 
                              param_expert = param_expert)

plot(ext_Surv, add.km = T, t = seq(0:180), ci = TRUE)


##TODO: how to get confidence intervals? make.surv()?

# obs_Surv <- survHE::fit.models(formula = Surv(death_t, death) ~ 1,
#                        data = dat_FCR,
#                        distr = "exponential",
#                        method = "hmc")

bs_Surv <- flexsurv::flexsurvreg(formula = Surv(time, event) ~ 1,
                       data = data_sim,
                       dist = "exp")

ble_Surv <- blendsurv(obs_Surv, ext_Surv, blend_interv, beta_params)

plot(ble_Surv)
```

Alternatively, we could use the `{expertsurv}` functions to fit both the data and the external information together as intended in the `{expertsurv}` package.
However, to modify this so that it is similar to `{blendr}` we can choose variance of constraint as functions of the blending interval.


TODO...
## Using sample size formula

```{r}
# external estimate
data_sim <- ext_surv_sim(t_info = 100,
                         S_info = 0.05,
                         T_max = 180,
                         n = 100)
                         # n = 40)
```

```{r}
ext_Surv3 <- flexsurv::flexsurvreg(formula = Surv(time, event) ~ 1,
                       data = data_sim,
                       dist = "exp")
```

```{r}
plot(ext_Surv3)
```

```{r}
summary(ext_Surv3)[[1]] |> head()
```

median time of mean 46
median time of upper limit 60
0.5 = exp(-lambda t) so lambda = log(2)/t
which gives log(2)/46 = 0.015 and log(2)/60 = 0.012
a hazard ratio of 46/60 = 0.77

```{r}
# log-mean method
# https://shariq-mohammed.github.io/files/cbsa2019/2-power-and-sample-size.html
expLogMeanDeaths = function(Delta, alpha, pwr){
  z.alpha = qnorm(alpha, lower.tail=FALSE)
  z.beta = qnorm(1-pwr, lower.tail=FALSE)
  num = (z.alpha + z.beta)^2
  denom = (log(Delta))^2

  num/denom
}
```

```{r}
# one-sided test
expLogMeanDeaths(Delta = 0.77, alpha = 0.025, pwr = 0.8)
```


